This analysis is organized into four sections: whole group, sex subgroup,
race subgroup, and power analysis. We use R [@R-base] and the
package `survival` [@survival-package] to perform the analysis and
generate exhibits in this paper.

```{r}
#| label: load-packages
#| echo: false
#| warning: false
#| message: false

library(GLDreg)
library(survival)
library(dplyr)
library(knitr)

qtable <- function(x, caption = NULL, align = NULL, row.names = FALSE, digits = 3) {
  knitr::kable(
    x,
    caption = caption,
    align = align,
    row.names = row.names,
    digits = digits,
    booktabs = knitr::is_latex_output()
  )
}

data(actg)
df <- actg

df$karnof <- as.numeric(as.character(df$karnof))

df$idv_label <- factor(
  df$tx,
  levels = c(0, 1),
  labels = c("No Indinavir", "Indinavir")
)

df$trt_label <- df$idv_label

df$sex_label <- factor(
  df$sex,
  levels = 0:1,
  labels = c("Female", "Male")
)

df$raceth_label <- factor(
  df$raceth,
  levels = 1:5,
  labels = c("White", "Black", "Hispanic", "Asian/PI", "Other")
)

df$race_white    <- as.integer(df$raceth == 1)
df$race_black    <- as.integer(df$raceth == 2)
df$race_hispanic <- as.integer(df$raceth == 3)
df$race_asian    <- as.integer(df$raceth == 4)
df$race_other    <- as.integer(df$raceth == 5)

n_total  <- nrow(df)
n_events <- sum(df$censor)
```

## Whole Group

We begin by replicating the primary finding from Hammer et al. (1997)
using the full study sample. The Kaplan-Meier curves in @fig-km-plot
show the probability of remaining event-free over time for each
treatment group. The two curves separate early and remain apart,
suggesting that indinavir reduced the hazard of AIDS progression or
death.

```{r}
#| label: fig-km-plot
#| fig-cap: "Kaplan-Meier survival curves for Indinavir versus No Indinavir treatment groups."
#| fig-pos: "!ht"
#| out-width: "80%"
#| echo: false
#| warning: false

km_fit <- survfit(Surv(time, censor) ~ idv_label, data = df)

plot(
  km_fit,
  col  = c("black", "steelblue"),
  lty  = 1,
  lwd  = 2,
  xlab = "Days",
  ylab = "Event-Free Survival Probability",
  main = "Kaplan-Meier Curves by Treatment Group"
)

legend(
  "topright",
  legend = c("No Indinavir", "Indinavir"),
  col    = c("black", "steelblue"),
  lty    = 1,
  lwd    = 2,
  bty    = "n",
  cex    = 0.85
)
```

To estimate the treatment effect while adjusting for patient
characteristics, we fit a Cox proportional hazards model. The model
includes treatment assignment, baseline CD4 count, age, Karnofsky
performance score, months of prior ZDV use, and CD4 count stratum. The
full regression results are shown in @tbl-whole-group.

```{r}
#| label: whole-group-model
#| echo: false
#| warning: false

model_baseline <- coxph(
  Surv(time, censor) ~ tx + age + cd4 + karnof + priorzdv + strat2,
  data = df,
  ties = "efron"
)

s        <- summary(model_baseline)
hr_whole <- exp(coef(model_baseline)["tx"])
ci_whole <- exp(confint(model_baseline)["tx", ])

logrank_whole <- survdiff(Surv(time, censor) ~ tx, data = df)
lr_chisq      <- round(logrank_whole$chisq, 3)
lr_p          <- round(1 - pchisq(logrank_whole$chisq, df = 1), 4)
```

```{r}
#| label: tbl-whole-group
#| tbl-cap: "Whole-Group Baseline Cox Model --- Hazard Ratios"
#| results: asis
#| echo: false
#| warning: false

coef_df <- data.frame(
  Covariate = rownames(s$coefficients),
  HR = round(exp(s$coefficients[, "coef"]), 3),
  `95% CI Lower` = round(exp(s$coefficients[, "coef"] - 1.96 * s$coefficients[, "se(coef)"]), 3),
  `95% CI Upper` = round(exp(s$coefficients[, "coef"] + 1.96 * s$coefficients[, "se(coef)"]), 3),
  `p-value` = round(s$coefficients[, "Pr(>|z|)"], 4),
  check.names = FALSE
)

coef_df$Covariate <- dplyr::recode(
  coef_df$Covariate,
  "tx"       = "Treatment: Indinavir vs. No Indinavir",
  "age"      = "Age",
  "cd4"      = "Baseline CD4",
  "karnof"   = "Karnofsky Score",
  "priorzdv" = "Prior ZDV Use",
  "strat2"   = "CD4 Stratum"
)

qtable(coef_df, caption = NULL, align = "lrrrr")
```

The estimated hazard ratio for treatment is HR =
`r round(hr_whole, 3)`. This means that at any point in time during
follow-up, patients receiving indinavir had approximately half the hazard
of experiencing an AIDS-defining event or death compared to the control
group, after adjusting for other covariates. This is consistent with the
original findings of Hammer et al. (1997). A log-rank test confirms
overall treatment group separation, with a chi-square statistic of
`r lr_chisq` on 1 degree of freedom (p = `r lr_p`), indicating that the
difference in event timing between the two treatment groups is
statistically significant.

## Sex Subgroup

To examine whether the treatment effect differed between men and women,
we fit two models. The first is a baseline model that includes sex as a
fixed covariate alongside treatment. The second is a full model that adds
a treatment-by-sex interaction term. A likelihood ratio test compares the
two models. We then fit separate Cox models for females and males to
estimate sex-specific hazard ratios.

```{r}
#| label: sex-models
#| echo: false
#| warning: false

model_sex_baseline <- coxph(
  Surv(time, censor) ~ tx + sex + age + cd4 + karnof + priorzdv + strat2,
  data = df,
  ties = "efron"
)

hr_sex_base <- exp(coef(model_sex_baseline)["tx"])
ci_sex_base <- exp(confint(model_sex_baseline)["tx", ])

model_sex_full <- coxph(
  Surv(time, censor) ~ tx * sex + age + cd4 + karnof + priorzdv + strat2,
  data = df,
  ties = "efron"
)

lr_sex     <- anova(model_sex_baseline, model_sex_full)
lr_s_chisq <- round(lr_sex[2, "Chisq"], 3)
lr_s_p     <- round(lr_sex[2, "Pr(>|Chi|)"], 4)

s_sb <- summary(model_sex_baseline)
s_sf <- summary(model_sex_full)

hr_sex_interaction <- round(exp(coef(model_sex_full)["tx:sex"]), 3)

sex_strata <- lapply(c(0, 1), function(g) {
  lbl <- ifelse(g == 0, "Female", "Male")
  sub <- df[df$sex == g, ]
  m <- coxph(
    Surv(time, censor) ~ tx + age + cd4 + karnof + priorzdv + strat2,
    data = sub,
    ties = "efron"
  )
  hr <- exp(coef(m)["tx"])
  ci <- exp(confint(m)["tx", ])
  data.frame(
    Subgroup = lbl,
    n = nrow(sub),
    Events = sum(sub$censor),
    HR = round(hr, 3),
    `95% CI Lower` = round(ci[1], 3),
    `95% CI Upper` = round(ci[2], 3),
    check.names = FALSE
  )
})

sex_strata_df <- do.call(rbind, sex_strata)

hr_female    <- sex_strata_df$HR[sex_strata_df$Subgroup == "Female"]
hr_male      <- sex_strata_df$HR[sex_strata_df$Subgroup == "Male"]
n_male       <- sex_strata_df$n[sex_strata_df$Subgroup == "Male"]
events_male  <- sex_strata_df$Events[sex_strata_df$Subgroup == "Male"]
```

The sex baseline and full interaction models are reported in Appendix
Tables A1 and A2. The baseline treatment hazard ratio remains
approximately unchanged at `r round(hr_sex_base, 3)`, and the coefficient
for sex is not statistically significant, indicating that sex does not
meaningfully confound the overall treatment effect. In the full
interaction model, the hazard ratio for treatment among females, the
reference group, is `r round(exp(coef(model_sex_full)["tx"]), 3)`. The
interaction hazard ratio is `r hr_sex_interaction`, but the likelihood
ratio test yields a p-value of `r lr_s_p`, so adding the interaction term
does not significantly improve model fit. We cannot conclude that sex
modifies the treatment effect.

The sex-specific models yield a hazard ratio of `r hr_female` for
female participants and `r hr_male` for male participants. With only
`r events_male` events among male participants across `r n_male` enrolled
individuals, the Cox model for males is severely underpowered, and the
estimated treatment effect should be interpreted with caution.

## Race/Ethnicity Subgroup

To examine whether the treatment effect differed across race and ethnicity groups, we fit a baseline model that includes race/ethnicity as a fixed covariate and a full model that adds treatment-by-race/ethnicity interaction terms. Race/ethnicity is coded as a five-level factor with White non-Hispanic participants as the reference group. We then fit separate Cox models within each race/ethnicity group to estimate group-specific hazard ratios.

```{r}
#| label: race-models
#| echo: false
#| warning: false

model_race_baseline <- coxph(
  Surv(time, censor) ~ tx + raceth_label + age + cd4 + karnof + priorzdv + strat2,
  data = df,
  ties = "efron"
)

hr_race_base <- exp(coef(model_race_baseline)["tx"])
ci_race_base <- exp(confint(model_race_baseline)["tx", ])

model_race_full <- coxph(
  Surv(time, censor) ~ tx * raceth_label + age + cd4 + karnof + priorzdv + strat2,
  data = df,
  ties = "efron"
)

lr_race    <- anova(model_race_baseline, model_race_full)
lr_r_chisq <- round(lr_race[2, "Chisq"], 3)
lr_r_p     <- round(lr_race[2, "Pr(>|Chi|)"], 4)

s_rf <- summary(model_race_full)

race_groups <- list(
  list(var = "race_white",    label = "White"),
  list(var = "race_black",    label = "Black"),
  list(var = "race_hispanic", label = "Hispanic"),
  list(var = "race_asian",    label = "Asian/PI"),
  list(var = "race_other",    label = "Other")
)

race_strata <- lapply(race_groups, function(g) {
  sub <- df[df[[g$var]] == 1, ]
  m <- coxph(
    Surv(time, censor) ~ tx + age + cd4 + karnof + priorzdv + strat2,
    data = sub,
    ties = "efron"
  )
  hr <- exp(coef(m)["tx"])
  ci <- exp(confint(m)["tx", ])
  data.frame(
    Subgroup = g$label,
    n = nrow(sub),
    Events = sum(sub$censor),
    HR = round(hr, 3),
    `95% CI Lower` = round(ci[1], 3),
    `95% CI Upper` = round(ci[2], 3),
    check.names = FALSE
  )
})

race_strata_df <- do.call(rbind, race_strata)
```

The baseline model yields a treatment hazard ratio of
`r round(hr_race_base, 3)`, indicating that after adjusting for
race/ethnicity, patients receiving indinavir had approximately 47% lower
hazard of AIDS-defining events or death. The full race/ethnicity
interaction model is reported in Appendix Table A3. The interaction terms
for Black, Hispanic, and Other are not statistically significant, providing no clear evidence that the treatment effect differs across these groups. The interaction term for the Asian/Pacific Islander group has a p-value of
`r round(s_rf$coefficients["tx:raceth_labelAsian/PI", "Pr(>|z|)"], 3)`,
but the confidence interval is extremely wide because there were only
`r race_strata_df$Events[race_strata_df$Subgroup == "Asian/PI"]` events
in that subgroup. The likelihood ratio test gives a chi-square statistic
of `r lr_r_chisq` with 4 degrees of freedom (p = `r lr_r_p`), so we
cannot conclude that race/ethnicity modifies the treatment effect.

## Summary of All Effect Sizes

@tbl-summary presents all hazard ratio estimates across the whole group,
sex, and race/ethnicity analyses. Estimates are stable for the overall
group and larger subgroups but become unreliable as event counts fall.
For Asian/PI and Other, the estimates are essentially uninformative. An
HR less than 1 indicates that indinavir-containing therapy reduces the
hazard of AIDS-defining events or death relative to nucleoside-only
therapy.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of Treatment Effect Sizes Across All Analyses"
#| results: asis
#| echo: false
#| warning: false

summary_rows <- list(
  data.frame(
    Analysis = "Whole Group", Model = "Baseline", Subgroup = "All",
    n = nrow(df), Events = sum(df$censor),
    HR = round(hr_whole, 3),
    Lower = round(ci_whole[1], 3),
    Upper = round(ci_whole[2], 3)
  ),
  data.frame(
    Analysis = "Sex", Model = "Baseline (pooled)", Subgroup = "All",
    n = nrow(df), Events = sum(df$censor),
    HR = round(hr_sex_base, 3),
    Lower = round(ci_sex_base[1], 3),
    Upper = round(ci_sex_base[2], 3)
  ),
  data.frame(
    Analysis = "Sex", Model = "Stratum-specific", Subgroup = "Female",
    n = sex_strata_df$n[1], Events = sex_strata_df$Events[1],
    HR = sex_strata_df$HR[1],
    Lower = sex_strata_df$`95% CI Lower`[1],
    Upper = sex_strata_df$`95% CI Upper`[1]
  ),
  data.frame(
    Analysis = "Sex", Model = "Stratum-specific", Subgroup = "Male",
    n = sex_strata_df$n[2], Events = sex_strata_df$Events[2],
    HR = sex_strata_df$HR[2],
    Lower = sex_strata_df$`95% CI Lower`[2],
    Upper = sex_strata_df$`95% CI Upper`[2]
  ),
  data.frame(
    Analysis = "Race", Model = "Baseline (pooled)", Subgroup = "All",
    n = nrow(df), Events = sum(df$censor),
    HR = round(hr_race_base, 3),
    Lower = round(ci_race_base[1], 3),
    Upper = round(ci_race_base[2], 3)
  ),
  data.frame(
    Analysis = "Race", Model = "Stratum-specific", Subgroup = "White",
    n = race_strata_df$n[1], Events = race_strata_df$Events[1],
    HR = race_strata_df$HR[1],
    Lower = race_strata_df$`95% CI Lower`[1],
    Upper = race_strata_df$`95% CI Upper`[1]
  ),
  data.frame(
    Analysis = "Race", Model = "Stratum-specific", Subgroup = "Black",
    n = race_strata_df$n[2], Events = race_strata_df$Events[2],
    HR = race_strata_df$HR[2],
    Lower = race_strata_df$`95% CI Lower`[2],
    Upper = race_strata_df$`95% CI Upper`[2]
  ),
  data.frame(
    Analysis = "Race", Model = "Stratum-specific", Subgroup = "Hispanic",
    n = race_strata_df$n[3], Events = race_strata_df$Events[3],
    HR = race_strata_df$HR[3],
    Lower = race_strata_df$`95% CI Lower`[3],
    Upper = race_strata_df$`95% CI Upper`[3]
  ),
  data.frame(
    Analysis = "Race", Model = "Stratum-specific", Subgroup = "Asian/PI",
    n = race_strata_df$n[4], Events = race_strata_df$Events[4],
    HR = race_strata_df$HR[4],
    Lower = race_strata_df$`95% CI Lower`[4],
    Upper = race_strata_df$`95% CI Upper`[4]
  ),
  data.frame(
    Analysis = "Race", Model = "Stratum-specific", Subgroup = "Other",
    n = race_strata_df$n[5], Events = race_strata_df$Events[5],
    HR = race_strata_df$HR[5],
    Lower = race_strata_df$`95% CI Lower`[5],
    Upper = race_strata_df$`95% CI Upper`[5]
  )
)

summary_df <- do.call(rbind, summary_rows)
colnames(summary_df)[7:8] <- c("95% CI Lower", "95% CI Upper")


qtable(summary_df, caption = NULL, align = "llllcccc", row.names = FALSE)


```

## Power Analysis

Before estimating sample sizes for future studies, we calculated the
observed power of the ACTG 320 analyses for the overall population and
each subgroup. Power was estimated assuming equal allocation between treatment and control groups. To account for multiple hypothesis testing, we used a Bonferroni-adjusted significance level of $\alpha = 0.025$ for each interaction test. This adjustment controls the family-wise error rate across the two primary subgroup comparisons (sex and race/ethnicity), ensuring that the overall probability of a Type I error remains bounded at 0.05.

```{r}
#| label: power-setup
#| echo: false
#| warning: false

alpha      <- 0.025
allocation <- 0.50

observed_power <- function(hr, events, alpha = 0.025, allocation = 0.50) {
  z_alpha <- qnorm(1 - alpha / 2)
  power <- pnorm(
    sqrt(events * allocation * (1 - allocation)) * abs(log(hr)) - z_alpha
  )
  round(power, 3)
}

required_events <- function(hr, alpha, power) {
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta  <- qnorm(power)
  ceiling((z_alpha + z_beta)^2 / (log(hr))^2)
}

required_n <- function(hr, alpha, power, event_rate) {
  ceiling(required_events(hr, alpha, power) / event_rate)
}

power_vals <- c(0.80, 0.85, 0.90, 0.95)

subgroups <- list(
  list(label = "Whole Group",    hr = 0.517, n = 1151, events = 96),
  list(label = "Sex: Female",    hr = 0.459, n = 951,  events = 81),
  list(label = "Race: White",    hr = 0.539, n = 596,  events = 51),
  list(label = "Race: Black",    hr = 0.287, n = 327,  events = 21),
  list(label = "Race: Hispanic", hr = 0.628, n = 203,  events = 20)
)
```

The observed power estimates are shown in @tbl-observed-power. Only the
overall group and the female subgroup exceeded the 80% power threshold.
All other subgroups fell below it, and for the male, Asian/Pacific
Islander, and Other subgroups, the underlying hazard ratio estimates are
themselves too unstable to produce meaningful power calculations.

```{r}
#| label: tbl-observed-power
#| tbl-cap: "Initial Observed Power by Overall Group and Subgroup"
#| results: asis
#| echo: false
#| warning: false

initial_power_df <- data.frame(
  Analysis = c(
    "Whole Group", "Sex", "Sex",
    "Race/Ethnicity", "Race/Ethnicity", "Race/Ethnicity",
    "Race/Ethnicity", "Race/Ethnicity"
  ),
  Subgroup = c(
    "All", "Female", "Male",
    "White", "Black", "Hispanic", "Asian/PI", "Other"
  ),
  n = c(
    nrow(df),
    sex_strata_df$n[sex_strata_df$Subgroup == "Female"],
    sex_strata_df$n[sex_strata_df$Subgroup == "Male"],
    race_strata_df$n[race_strata_df$Subgroup == "White"],
    race_strata_df$n[race_strata_df$Subgroup == "Black"],
    race_strata_df$n[race_strata_df$Subgroup == "Hispanic"],
    race_strata_df$n[race_strata_df$Subgroup == "Asian/PI"],
    race_strata_df$n[race_strata_df$Subgroup == "Other"]
  ),
  Events = c(
    sum(df$censor),
    sex_strata_df$Events[sex_strata_df$Subgroup == "Female"],
    sex_strata_df$Events[sex_strata_df$Subgroup == "Male"],
    race_strata_df$Events[race_strata_df$Subgroup == "White"],
    race_strata_df$Events[race_strata_df$Subgroup == "Black"],
    race_strata_df$Events[race_strata_df$Subgroup == "Hispanic"],
    race_strata_df$Events[race_strata_df$Subgroup == "Asian/PI"],
    race_strata_df$Events[race_strata_df$Subgroup == "Other"]
  ),
  `Observed HR` = c(
    round(hr_whole, 3),
    sex_strata_df$HR[sex_strata_df$Subgroup == "Female"],
    sex_strata_df$HR[sex_strata_df$Subgroup == "Male"],
    race_strata_df$HR[race_strata_df$Subgroup == "White"],
    race_strata_df$HR[race_strata_df$Subgroup == "Black"],
    race_strata_df$HR[race_strata_df$Subgroup == "Hispanic"],
    race_strata_df$HR[race_strata_df$Subgroup == "Asian/PI"],
    race_strata_df$HR[race_strata_df$Subgroup == "Other"]
  ),
  check.names = FALSE
)

initial_power_df$`Event Rate` <- round(
  initial_power_df$Events / initial_power_df$n,
  3
)

initial_power_df$`Observed Power` <- mapply(
  observed_power,
  hr = initial_power_df$`Observed HR`,
  events = initial_power_df$Events,
  MoreArgs = list(alpha = alpha, allocation = allocation)
)

initial_power_df$Interpretation <- ifelse(
  initial_power_df$`Observed Power` >= 0.80,
  "At or above 80%",
  "Below 80%"
)



qtable(initial_power_df, caption = NULL, align = "llcccccl", row.names = FALSE)


```

Detailed sample size calculations are provided in Appendix Tables A4--A6.
For subgroups with more stable hazard ratio estimates, required sample
sizes were calculated at 80%, 85%, 90%, and 95% power. For male,
Asian/PI, and Other subgroups, where the observed hazard ratios were
unstable, required sample sizes were calculated across assumed hazard
ratios from 0.9 to 1.1. The Hispanic subgroup stands out as the only
stable subgroup that was underpowered in ACTG 320, requiring
approximately `r required_n(0.628, 0.025, 0.80, 20 / 203)` participants
to achieve 80% power compared to the `r 203` enrolled.

```{r}
#| label: fig-power-curves
#| fig-cap: "Power curves showing achieved power as a function of sample size for each reliable subgroup, with the 80% power threshold shown."
#| echo: false
#| warning: false

n_seq  <- seq(50, 3000, by = 25)
cols   <- c("black", "steelblue", "tomato", "forestgreen", "purple")
labels <- sapply(subgroups, function(sg) sg$label)

plot(
  NULL,
  xlim = c(50, 3000),
  ylim = c(0, 1),
  xlab = "Total Sample Size (n)",
  ylab = "Statistical Power",
  main = "Power Curves by Subgroup"
)

abline(h = 0.80, lty = 2, col = "gray40")
text(2800, 0.82, "80% Power", cex = 0.75, col = "gray40")

for (i in seq_along(subgroups)) {
  sg <- subgroups[[i]]
  event_rate <- sg$events / sg$n
  z_alpha <- qnorm(1 - alpha / 2)

  pwr <- sapply(n_seq, function(n) {
    d <- n * event_rate
    se <- 1 / sqrt(d * 0.25)
    ncp <- abs(log(sg$hr)) / se
    pnorm(ncp - z_alpha) + pnorm(-ncp - z_alpha)
  })

  lines(n_seq, pwr, col = cols[i], lwd = 2)
}

legend(
  "bottomright",
  legend = labels,
  col = cols,
  lwd = 2,
  bty = "n",
  cex = 0.8
)
```

```{r}
#| label: tbl-power-adequacy
#| tbl-cap: "Adequacy of ACTG 320 Sample Sizes for 80% Power"
#| results: asis
#| echo: false
#| warning: false
#| message: false

rec_rows <- lapply(subgroups, function(sg) {
  event_rate <- sg$events / sg$n
  n80 <- required_n(sg$hr, alpha, 0.80, event_rate)
  adequacy <- ifelse(sg$n >= n80, "Adequate", "Underpowered")
  data.frame(
    Subgroup  = sg$label,
    `Observed HR` = sg$hr,
    `ACTG 320 n` = sg$n,
    `Required n for 80% Power` = n80,
    `ACTG 320 Status` = adequacy,
    check.names = FALSE
  )
})

rec_df <- do.call(rbind, rec_rows)

qtable(rec_df, caption = NULL, align = "lcccl", row.names = FALSE)
```